\bf Eye Fitting Straight Lines in the Modern Era

### 2022-01-27 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
proof read of manuscript.

We all use statistical graphics, but how do we know that the graphics we
use are communicating properly? When creating a graphic, we must
consider the design choices most effective for conveying the intended
result. For instance, we must decide to highlight the relationship
between two variables in a scatterplot by including a trend line, or
adding color to highlight clustering \citep{vanderplas2017clusters}.
These design choices require that we understand the perception and
visual biases that come into play when creating graphics, and as
graphics are evaluated visually, we must use human testing to ground our
understanding in empiricism.

@@

In \citet{ciccione2021can}, participants were asked to judge trends,
estimate slopes, and conduct extrapolation. To estimate slopes,
participants were asked to report the slope of the best-fitting
regression line using a trackpad to adjust the tilt of a line on screen.
Results indicated the slopes participants reported were always in excess
of the ideal slopes, both in the positive and in the negative direction,
and those biases increase with noise and with number of points. This
supports the results found in \citet{mosteller1981eye} and suggest that
participants might use Deming regression when fitting a line to a noisy
scatterplot.

@@

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as a tool for graphical testing. The `You Draw It' method
is validated by replicating the study conducted by
\citet{mosteller1981eye}. Based on previous research surrounding
``ensemble perception,'' we hypothesize that visual regression tends to
mimic principle component regression rather than a ordinary least
squares regression. In order to assess this hypothesis, we introduce a
method for statistically modeling the participant drawn lines using
generalized additive mixed models.

@@

Data Driven Documents (D3), a JavaScript-based graphing framework that
facilitates user interaction, is used to create the `You Draw It' task
plots. Integrating this into RShiny using the \texttt{r2d3} package,
participants are asked to draw a trend-line using their computer mouse
through a scatter-plot shown on their screen. In the study, participants
are shown an interactive scatter-plot (\cref{fig:ydi-stimuli}) along
with the prompt, ``Use your mouse to fill in the trend in the yellow box
region.'' The yellow box region moves along as the participant draws
their trend-line until the yellow region disappears, indicating the
participant has filled in the entire domain. Details of the development
of the `You Draw It' task plots will be addressed in future work.

@@

Model equation parameters, \(\beta_0\) and \(\beta_1\), were selected to
reflect the four data sets (F, N, S, and V) used in
\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). Parameter
choices F, N, and S simulated data across a domain of 0 to 20. Parameter
choice F produces a trend with a positive slope and a large variance
while N has a negative slope and a large variance. In comparison, S
shows a trend with a positive slope with a small variance and V yields a
steep positive slope with a small variance over the domain of 4 to 16.
\cref{fig:eyefitting-simplot} illustrates an example of simulated data
for all four parameter choices intended to reflect the trends in
\citet{mosteller1981eye}. Aesthetic design choices were made consistent
across each of the interactive `You Draw It' task plots. The y-axis
range extended 10\% beyond (above and below) the range of the simulated
data points to allow for users to draw outside the simulated data set
range.

@@

This experiment was conducted as part of a larger study; for simplicity,
we focus on the study design and methods related to the current study.
Each scatter-plot was the graphical representation of a data set that
was generated randomly, independently for each participant at the start
of the experiment. Participants in the study are shown two `You Draw It'
practice plots in order to train participants in the skills associated
with executing the task. During the practice session, participants are
provided with instruction prompts accompanied by a .gif and a practice
plot. Instructions guide participants to start at the edge of the yellow
box, to make sure the yellow region is moving along with their mouse as
they draw, and that they can draw over their already drawn line.
Practice plots are then followed by four `You Draw It' task plots
associated with the current study. The order of the task plots was
randomly assigned for each individual in a completely randomized design.

@@

We compare the participant drawn line to two regression lines determined
by ordinary least squares regression and regression based on the
principal axis (i.e.~Deming Regression). \cref{fig:ols-vs-pca-example}
illustrates the difference between an OLS regression line which
minimizes the vertical distance of points from the line and a regression
line based on the principal axis which minimizes the Euclidean distance
of points (orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope
of the linear regression line fit through the simulated points could
differ from the predetermined slope. Therefore, we fit an ordinary least
squares (OLS) regression to each scatter-plot to obtain estimated
parameters \(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted
values, \(\hat y_{k,OLS}\), are then obtained every 0.25 increments
across the domain from the OLS regression equation,
\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\)., for
\(k = 1, ..., 4 x_{max} +1\). The regression equation based on the
principal axis was determined by using the \texttt{princomp} function in
the stats package in base R to obtain the rotation of the coordinate
axes from the first principal component (direction which captures the
most variance). The estimated slope, \(\hat\beta_{1,PCA}\), is
determined by the ratio of the axis rotation in y and axis rotation in x
of the first principal component with the y-intercept,
\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
Fitted values, \(\hat y_{k,PCA}\), are then obtained every 0.25
increment across the domain from the PCA regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).

@@

\caption{Illustrates the data associated with and collected for one 'You Draw It' task plot. Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  x-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
  fit
\item
  \(\gamma_0\) is the overall intercept
\item
  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
  V, N) on the intercept
\item
  \(\gamma_1\) is the overall slope for \(x\)
\item
  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
\item
  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(j^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

Constraining the residual trend to a linear fit,
\cref{fig:eyefitting-lmer-residualplots} shows the estimated trend line
of the residuals between the participant drawn points and fitted values
for both the OLS regression line and PCA regression line. Estimated
residual trend lines are overlaid on the observed individual participant
residuals. Results indicate the estimated trends of PCA residuals
(orange) appear to align closer to the \(y=0\) horizontal (dashed) line
than the OLS residuals (blue). In particular, this trend is more
prominent in parameter choices with large variances (F and N). These
results are consistent to those found in \citet{mosteller1981eye}
indicating participants fit a trend-line closer to the estimated
regression line with the slope of based on the first principal axis than
the estimated OLS regression line thus, providing support for ``ensemble
perception''.

@@

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
  choice, \(j^{th}\) participant, and \(k^{th}\) increment of x-value
\item
  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  x-value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
  \(k^{th}\) increment of x-value corresponding to either the OLS or PCA
  fit
\item
  \(\alpha_i\) is the intercept for the parameter choice \(i\)
\item
  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
\item
  \(x_{ijk}\) is the x-value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the error due to
  participant variation
\item
  \(s_{j}\) is the random smoothing spline for each participant.
\end{itemize}

Allowing for flexibility in the residual trend,
\cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line
of the residuals between the participant drawn points and fitted values
for both the OLS regression line and PCA regression line. Estimated
residual trends are overlaid on the observed individual participant
residuals. The results of the GAMM align with those shown in
\cref{fig:eyefitting-lmer-residualplots} providing support that for
scatter-plots with more noise (F and N), estimated trends of PCA
residuals (orange) appear to align closer to the \(y=0\) horizontal
(dashed) line than the OLS residuals (blue). By fitting smoothing
splines, we can determine whether participants naturally fit a straight
trend-line to the set of points or whether they deviate throughout the
domain. In particular, in scatter-plots with smaller variance (S and V),
we can see that participants began at approximately the correct starting
point then deviated away from the fitted regression lines and correcting
for their fit toward the end of their trend-line. In scatter-plots with
larger variance (F and N), participants estimated their starting value
in the extreme direction of the OLS regression line based on the
increasing or decreasing trend but more accurately represented the
starting value of the PCA regression line. As participants continued
their trend-line, they crossed through the OLS regression line
indicating they estimated the slope in the extreme direction. These
results provide further insight into the curvature humans perceive in a
set of points.

@@

The intent of research was to adapt `You Draw It' from the New York
Times feature as a tool and method for testing graphics and introduce a
method for statistically modeling the participant drawn lines. We
provided support for the validity of the `You Draw It' method by
replicating the study found in \citet{mosteller1981eye}. Using
generalized additive mixed models, we assessed the deviation of the
participant drawn lines from the statistically fitted regression lines.
Our results found that when shown points following a linear trend,
participants visually fit a regression line that mimics the first
principle component regression as opposed to ordinary least squares
regression. Data simulated with a larger variance provided strong
support for a participants tendency to visually fit the first principle
component regression. Our results indicate that participants minimized
the distance from the their regression line over both the \(x\) and
\(y\) axis simultaneously. These results provide support that humans
perform ``ensemble perception'' in a statistical graphic setting. We
allowed participants to draw trend lines that deviated from a straight
line and gained an insight into the curvature the human eye perceives in
a set of points.

@@

This study provided a basis for the use of `You Draw It' as a tool for
testing statistical graphics and introduced a method for statistically
modeling participant drawn lines using generalized additive mixed
models. Further investigation is necessary to implement this method in
non-linear settings and with real data. This tool could also be used to
evaluate human ability to extrapolate data from trends. In the future,
an R package designed for easy implementation of `You Draw It' task
plots will help make this tool accessible to other researchers.

>>>>

We all use statistical graphics, but how do we know that the graphics we
use are communicating properly? When creating a graphic, we must
consider the design choices most effective for conveying the intended
result. For instance, we must decide to highlight the relationship
between two variables in a scatter-plot by including a trend line, or
adding color to highlight clustering \citep{vanderplas2017clusters}.
These design choices require that we understand the perception and
visual biases that come into play when creating graphics, and as
graphics are evaluated visually, we must use human testing to ground our
understanding in empiricism.

@@

In \citet{ciccione2021can}, participants were asked to judge trends,
estimate slopes, and conduct extrapolation. To estimate slopes,
participants were asked to report the slope of the best-fitting
regression line using a track-pad to adjust the tilt of a line on
screen. Results indicated the slopes participants reported were always
in excess of the ideal slopes, both in the positive and in the negative
direction, and those biases increase with noise and with number of
points. This supports the results found in \citet{mosteller1981eye} and
suggest that participants might use Deming regression when fitting a
line to a noisy scatter-plot.

@@

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as a tool for graphical testing. The `You Draw It' method
is validated by replicating the study conducted by
\citet{mosteller1981eye}. Based on previous research surrounding
``ensemble perception,'' we hypothesize that visual regression tends to
mimic principle component regression rather than an ordinary least
squares regression. In order to assess this hypothesis, we introduce a
method for statistically modeling the participant drawn lines using
generalized additive mixed models.

@@

In the study, participants are shown an interactive scatter-plot
(\cref{fig:ydi-stimuli}) along with the prompt, ``Use your mouse to fill
in the trend in the yellow box region.'' The yellow box region moves
along as the user draws their trend-line, providing a visual cue which
indicates where the user still needs to complete a trend line. After the
entire domain has been visually estimated or predicted, the yellow
region disappears, indicating the participant has completed the task.
Data Driven Documents (D3), a JavaScript-based graphing framework that
facilitates user interaction, is used to create the `You Draw It'
visual. In order to allow for user interaction and data collection, we
integrate the D3 visual into RShiny using the \texttt{r2d3} package.
Details of the development of the `You Draw It' task plots will be
addressed in future work.

@@

Model equation parameters, \(\beta_0\) and \(\beta_1\), were selected to
reflect the four data sets (F, N, S, and V) used in
\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). We obtain
\(\beta_0\) from the point-slope equation of a line using the mean of
the generated \(x\) values and the predefined \(y\) value at \(\bar x\),
denoted \(y_{\bar x}\). Parameter choices F, N, and S simulated data
across a domain of 0 to 20. Parameter choice F produces a trend with a
positive slope and a large variance while N has a negative slope and a
large variance. In comparison, S shows a trend with a positive slope
with a small variance and V yields a steep positive slope with a small
variance over the domain of 4 to 16. \cref{fig:eyefitting-simplot}
illustrates an example of simulated data for all four parameter choices
intended to reflect the trends in \citet{mosteller1981eye}. Aesthetic
design choices were made consistent across each of the interactive `You
Draw It' task plots. The y-axis range extended 10\% beyond (above and
below) the range of the simulated data points to allow for users to draw
outside the simulated data set range and avoid anchoring their lines to
the corners of the plot.

@@

This experiment was conducted as part of a larger study; for simplicity,
we focus on the study design and methods related to the current study.
Each scatter-plot was the graphical representation of a data set that
was generated randomly and independently for each participant at the
start of the experiment. Participants in the study are shown two `You
Draw It' practice plots in order to train participants in the skills
associated with executing the task. During the practice session,
participants are provided with instruction prompts accompanied by a .gif
and a practice plot. Instructions guide participants to start at the
edge of the yellow box, to make sure the yellow region is moving along
with their mouse as they draw, and that they can draw over their already
drawn line. Practice plots are then followed by four `You Draw It' task
plots associated with the current study. The order of the task plots was
randomly assigned for each individual in a completely randomized design.

@@

We compare the participant drawn line to two regression lines determined
by ordinary least squares (OLS) regression and regression based on the
principal axis (PCA). \cref{fig:ols-vs-pca-example} illustrates the
difference between an OLS regression line which minimizes the vertical
distance of points from the line and a regression line based on the
principal axis which minimizes the Euclidean distance of points
(orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope
of the linear regression line fit through the simulated points could
differ from the predetermined slope. Therefore, we fit an OLS regression
to each scatter-plot to obtain estimated parameters
\(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted values,
\(\hat y_{k,OLS}\), are then obtained every 0.25 increments across the
domain from the OLS regression equation,
\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\)., for
\(k = 1, ..., 4 x_{max} +1\). The regression equation based on the
principal axis was determined by using the \texttt{princomp} function in
the stats package in base R to obtain the rotation of the coordinate
axes from the first principal component (direction which captures the
most variance). The estimated slope, \(\hat\beta_{1,PCA}\), is
determined by the ratio of the axis rotation in \(y\) and axis rotation
in \(x\) of the first principal component with the y-intercept,
\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
Fitted values, \(\hat y_{k,PCA}\), are then obtained every 0.25
increment across the domain from the PCA regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).

@@

\caption{Illustrates the data associated with and collected for one `You Draw It' task plot. Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value
\item
  \(\hat y_{ijk,fit}\) is the fitted \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted \(y\)
  values for the \(i^{th}\) parameter choice, \(j^{th}\) participant,
  and \(k^{th}\) increment of \(x\) value corresponding to either the
  OLS or PCA fit
\item
  \(\gamma_0\) is the overall intercept
\item
  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
  V, N) on the intercept
\item
  \(\gamma_1\) is the overall slope for \(x\)
\item
  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
\item
  \(x_{ijk}\) is the \(x\) value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the random error due to
  the \(j^{th}\) participant's characteristics
\item
  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
\end{itemize}

Constraining the residual trend to a linear fit,
\cref{fig:eyefitting-lmer-residualplots} shows the estimated trend line
of the residuals between the participant drawn points and fitted values
for both the OLS regression line and PCA regression line. Estimated
residual trend lines are overlaid on the observed individual participant
residuals. Results indicate the estimated trends of PCA residuals
(orange) appear to align closer to the \(y=0\) horizontal (dashed) line
than the OLS residuals (blue). In particular, this trend is more
prominent in parameter choices with large variances (F and N). These
results are consistent to those found in \citet{mosteller1981eye}
indicating participants fit a trend-line closer to the estimated
regression line with a slope based on the first principal axis than the
estimated OLS regression line thus, providing support for ``ensemble
perception''.

@@

\begin{itemize}
\tightlist
\item
  \(y_{ijk,drawn}\) is the drawn \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value
\item
  \(\hat y_{ijk,fit}\) is the fitted \(y\) value for the \(i^{th}\)
  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
  \(x\) value corresponding to either the OLS or PCA fit
\item
  \(e_{ijk,fit}\) is the residual between the drawn and fitted \(y\)
  values for the \(i^{th}\) parameter choice, \(j^{th}\) participant,
  and \(k^{th}\) increment of \(x\) value corresponding to either the
  OLS or PCA fit
\item
  \(\alpha_i\) is the intercept for the parameter choice \(i\)
\item
  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
\item
  \(x_{ijk}\) is the \(x\) value for the \(i^{th}\) parameter choice,
  \(j^{th}\) participant, and \(k^{th}\) increment
\item
  \(p_{j} \sim N(0, \sigma^2_{participant})\) is the error due to
  participant variation
\item
  \(s_{j}\) is the random smoothing spline for each participant.
\end{itemize}

Allowing for flexibility in the residual trend,
\cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line
of the residuals between the participant drawn points and fitted values
for both the OLS regression line and PCA regression line. Estimated
residual trends are overlaid on the observed individual participant
residuals. The results of the GAMM align with those shown in
\cref{fig:eyefitting-lmer-residualplots} providing support that for
scatter-plots with more noise (F and N), estimated trends of PCA
residuals (orange) appear to align closer to the \(y=0\) horizontal
(dashed) line than the OLS residuals (blue). By fitting smoothing
splines, we can determine whether participants naturally fit a straight
trend-line to the set of points or whether they deviate throughout the
domain. In particular, in scatter-plots with smaller variance (S and V),
we can see that participants began at approximately the correct starting
point then deviated away from the fitted regression lines and corrected
for their fit toward the end of their trend-line. In scatter-plots with
larger variance (F and N), participants estimated their starting value
in the extreme direction of the OLS regression line based on the
increasing or decreasing trend but more accurately represented the
starting value of the PCA regression line. As participants continued
their trend-line, they crossed through the OLS regression line
indicating they estimated the slope in the extreme direction. These
results provide further insight into the curvature humans perceive in a
set of points.

@@

The intent of this research was to adapt `You Draw It' from the New York
Times feature as a tool and method for testing graphics and introduce a
method for statistically modeling the participant drawn lines. We
provided support for the validity of the `You Draw It' method by
replicating the study found in \citet{mosteller1981eye}. Using
generalized additive mixed models, we assessed the deviation of the
participant drawn lines from the statistically fitted regression lines.
Our results found that when shown points following a linear trend,
participants visually fit a regression line that mimics the first
principle component regression as opposed to ordinary least squares
regression. Data simulated with a larger variance provided strong
support for a participants tendency to visually fit the first principle
component regression. Our results indicate that participants minimized
the distance from the their regression line over both the \(x\) and
\(y\) axis simultaneously. These results provide support that humans
perform ``ensemble perception'' in a statistical graphic setting. We
allowed participants to draw trend lines that deviated from a straight
line and gained an insight into the curvature the human eye perceives in
a set of points.

@@

This study provided a basis for the use of `You Draw It' as a tool for
testing statistical graphics and introduced a method for statistically
modeling participant drawn lines using generalized additive mixed
models. Further investigation is necessary to implement this method in
non-linear settings and with real data in order to facilitate scientific
communication. This tool could also be used to evaluate human ability to
extrapolate data from trends. We intend to create an R package designed
for easy implementation of `You Draw It' task plots in order to make
this tool accessible to other researchers.
